---
    title: "Human Activity Recognition"
    author: "Susana Arias Laso"
    output: html_document
---

## Introduction

The goal of this project is to implement a machine learning algorithm to
predict how well a particular exercise was performed using data collected
at different locations of an accelerometer. The "classe" feature in the
dataset indicates the manner in which the exercise was done. Class A indicates
that the exercise was performed exactly according to the specification, while
the other four classes indicate some common mistake. Using accelerometers 
located on the belt, forearm, arm and dumbbell as recording devices, data from 
6 participants was collected in order to quantify how well a particular 
activity was done.

The following libraries are needed to run the code. 
```{r, results='hide', message=FALSE, warning=FALSE}
library(caret)
library(ggplot2)
library(dplyr)
library(magrittr)
library(tidyr)
library(doMC)

library(GGally)
```

Additionally, a seed is set before processing the data so the results can be 
easily reproduced.
```{r, echo=TRUE}
set.seed(2195324)
```

In order to save running time when implementing the prediction model, parallel 
processing is applied using half of the cores available.
```{r, echo=TRUE}
numCores <- detectCores()
registerDoMC(numCores / 2)
```

## Data Processing

```{r, echo=TRUE}
train_url = 
    'https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv'
if (!file.exists("training.csv")) {
    download.file(url = train_url, destfile = 'training.csv')
}
```

Some missing entries in the dataset were identified and considered "NA" values
before processing the data into training and test sets.

```{r, echo=TRUE}
training_data = read.csv(file = 'training.csv', 
                         na.strings = c("NA", "", "#DIV/0!"))
```

In order to deal with the missing values, a boolean vector with the number of 
elements corresponding to the number of features in the dataset was created. 
Each component in the vector corresponds to the total number of NA values of a
particular feature.

```{r, echo=TRUE}
na_col <- function(column_indx){
    sum(is.na(training_data[, column_indx]))
}

boolean <- function(indx){
    na_vector[indx] == 0
}
```

```{r, echo=TRUE}
indx_vector = 1:ncol(training_data)
na_vector = sapply(indx_vector, na_col)
boolean_vector = sapply(indx_vector, boolean)
```

The dataset in then evaluated on that boolean vector in order to keep only the
features without missing values.
```{r, echo=TRUE}
data = training_data[, boolean_vector]
```

As features of interest for the prediction model we will keep only the 
variables that contain the original sensor readings. This corresponds to 
selecting the features that contain the Euler angles (roll, pitch and yaw)
and the 3-axis coordinates of the accelerometer, gyroscope, and magnetometer
readings for each sensor. 

```{r, echo=TRUE, warning=FALSE}
data <- data %>%
    select(-contains("user"), -contains("timestamp"), -contains("window"), 
           -contains("total")) %>%
    select(-starts_with("X")) %>%
    mutate("classe" = as.factor(classe))
```

The data is then separated into training and test sets, using 70% of the
dataset to train the model.
```{r, echo=TRUE}
inTrain = createDataPartition(y = data$classe, p = 0.7, list = FALSE)
training = data[inTrain, ]
testing = data[-inTrain, ]
```

## Exploratory Data Analysis
Using the GGally library the correlation among the arm Euler angles is 
explored. 
```{r, echo=TRUE}
data_pairs <- select(training,
                     contains("roll_arm"), 
                     contains("pitch_arm"), 
                     contains("yaw_arm"), 
                     contains("classe"))
```

```{r, echo=TRUE}
ggallyDensityPlot <- function(data, mapping, ...){
    ggplot(data = data, mapping = mapping) + 
    geom_density(mapping = aes_string(color = "classe"), fill = NA)
}

ggpairs(data_pairs, columns = c(1:3), 
        mapping = ggplot2::aes(colour = classe),
        lower = list(continuous = wrap("points", alpha = 0.5)),
        diag = list(continuous = ggallyDensityPlot),
        title = 'Correlation among arm Euler angles' 
        )
```

No direct correlation is evident among these features, as one can notice in the
lower diagonal plots. The density plots suggest that the observations follow
a Gaussian distribution.

## Prediction Model

A random forest model was trained in order to predict how well each exercise
was performed, i.e., the "classe" variable. A 10-fold cross validation of the 
model was implemented, using a tuning parameter of tuneLength = 8 which 
indicates the number of values to try for each hyperparameter. The process of 
running the model took about 40 minutes.

```{r, echo=TRUE}
rf_file = "./rf10foldCV.rds"

if (file.exists(rf_file)){
    rf10fold_cv <- readRDS("./rf10foldCV.rds")
} else {
    rf10fold_cv <- train(classe ~ ., data = training, 
                    method = 'rf', prox = TRUE, 
                    trControl = trainControl(method = "cv", number = 10), 
                    tuneLength = 8)
}
```

The predictions of the model are evaluated on the test set, and shown in the
form of an accuracy table, where the number of cases predicted correctly are
indicated along the diagonal.

```{r, echo=TRUE}
pred <- predict(rf10fold_cv, testing)
accuracy_table <- table(pred, testing$classe)
accuracy_table
```

The model was 99.5% accurate on the test set, which indicates it will predict
the feature "classe" with highly enough precision on unseen data.
```{r, echo=TRUE}
confusionMatrix(pred, testing$classe)$overall[1]
```

The precision with which each class was predicted is listed below.
```{r, echo=TRUE}
acc <- function(index){
    accuracy_table[index, index] / sum(accuracy_table[, index])
}

classes <- c(1:5)
pred_by_class <- sapply(classes, acc)

for (i in seq(classes)) {
    cat("Class", classes[i], "precision: ", pred_by_class[i], "\n")
}
```

## Prediction Quiz

The following code applied the machine learning algorithm from the previous
section on 20 different cases available in the testing data. The predictions
are printed below.

```{r, echo=TRUE}
test_url = 
    'https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv'

if (!file.exists("test.csv")) {
    download.file(url = test_url, destfile = 'test.csv')
}

test_data = read.csv(file = 'test.csv', na.strings = c("NA", "", "#DIV/0!"))
```

```{r, echo=TRUE}
rf_predictions <- predict(rf10fold_cv, newdata = test_data)
rf_predictions
```






















































